---
title: "Use tKOI with an Agent and Its Analysis Graph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use tKOI with an Agent and Its Analysis Graph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Prepare an Agent Workflow

[tKOIAgent](https://github.com/BaranziniLab/tKOIAgent) version 2 provides the
`tkoi-agent` plugin. It combines two skills with an MCP server that reads a local
tKOI analysis:

* `tkoi-analysis` guides input preparation and runs the R package.
* `tkoi-knowledge-graph` searches nodes, follows connections, and reads enrichment
  results from the graph saved with that analysis.

For Codex, Claude Code, or BioRouter, follow the plugin's
[README](https://github.com/BaranziniLab/tKOIAgent/blob/main/README.md) and
[setup guide](https://github.com/BaranziniLab/tKOIAgent/blob/main/skills/tkoi-analysis/references/setup.md)
to install and configure the plugin. Use its
[preprocessing guide](https://github.com/BaranziniLab/tKOIAgent/blob/main/skills/tkoi-analysis/references/preprocessing.md)
to prepare gene identifiers, fold changes, and p-values before analysis. The R
package requires `gene_name` (Ensembl gene ID), `logfc`, and `pvalue` columns.

Ask the agent to use `tkoi-analysis` to check and prepare your input, run tKOI,
and save the analysis and provenance. The graph workflow requires tkoi 1.3.0 or
later so the saved result contains the exact graph used for the run.

# Keep the Analysis and Graph Together

`run_tkoi()` stores the igraph supplied as `subnetwork` in `result@subnetwork`.
This includes its vertex and edge attributes and applies to both the package's
default graph and custom graphs. `get_analysis_graph()` retrieves that stored
graph without substituting a different network.

After running an analysis in R, save the complete result:

```{r save-analysis, eval=FALSE}
graph = tkoi::get_analysis_graph(result)
saveRDS(result, "analysis.rds")
analysis_path = normalizePath("analysis.rds", mustWork = TRUE)
analysis_path
```

Give the plugin's `connect_analysis` tool the absolute file path as its `path`
argument, for example:

```json
{"path": "/absolute/path/to/analysis.rds"}
```

The local connection reads the graph and enrichment tables from this result.
`connect_analysis` returns an `analysis_id`, the saved file's MD5 identifier.
Keep that value and supply it as the first argument to every subsequent MCP
tool call. This prevents another agent's connection from silently changing the
graph used by your queries. The server rejects stale IDs or changed files; if
you save a revised result, reconnect explicitly and use its returned ID.

An older result without its stored graph cannot establish which graph produced
the statistics. `get_analysis_graph()` and the plugin reject such a result;
rerun the analysis with its original graph and save a new result. The package's
current `tkoi_net` is not used as a fallback.

# Explore with the MCP Tools

Connect the result first, then inspect its schema and resolve node IDs before
traversing. Every query names the intended analysis through `analysis_id`.

| Tool | Use |
| --- | --- |
| `connect_analysis(path)` | Open the local RDS result and its saved graph; return `analysis_id`. |
| `get_graph_schema(analysis_id)` | Inspect node types, vertex and edge attributes, and graph direction. |
| `search_nodes(analysis_id, query, node_type = NULL, limit = 25)` | Find labels or identifiers and obtain the graph's node IDs. |
| `get_node_neighbors(analysis_id, node_id, hops = 1, limit = 100)` | Inspect nearby nodes within a bounded number of hops. |
| `get_path_between_nodes(analysis_id, source_id, target_id, max_hops = 3)` | Find a connecting path within the requested hop limit. |
| `get_enrichment_results(analysis_id, node_type, limit = 25)` | Read a node type's enrichment table from the specified result. |

For example, ask the agent to inspect the schema, list enriched biological
processes, resolve a gene and process to node IDs, and examine their connections.
Use IDs returned by the tools: a gene symbol or Ensembl identifier is not
necessarily the graph's vertex name. The
[knowledge graph skill](https://github.com/BaranziniLab/tKOIAgent/blob/main/skills/tkoi-knowledge-graph/SKILL.md)
describes the exploration workflow in detail.

# Use the Same Graph Directly in R

The same exploration can be done without an agent. Read the saved result, retrieve
its graph, and pass that graph explicitly to helper functions:

```{r direct-r, eval=FALSE}
result = readRDS("analysis.rds")
graph = tkoi::get_analysis_graph(result)
node_id = result@network_summary_statistics$BiologicalProcess$node_id[1]

neighbors = tkoi::get_neighboring_nodes(
  node_id,
  degree_expansion = 1,
  subnetwork = graph
)

igraph::vertex_attr_names(graph)
igraph::edge_attr_names(graph)

# plot_network() defaults to the graph stored in result
tkoi::plot_network(result, target_node_id = node_id, degree_expansion = 2)
```

# Interpret the Connections

The bundled graph is undirected. Its vertex attributes include `name` (an opaque
node ID), `identifier`, `source`, `labels`, and `degree`. Its edges store
`edge_type`, for example `ISA_AiA` or `PARTOF_ApA`. These labels do not establish
direction or an activating or inhibiting effect. The analysis and graph traversal
treat connections in both directions.

Report only relation labels and provenance present in the actual graph
attributes. A vertex's `source` attribute is not evidence for an edge. The
bundled edge attributes do not supply evidence records or source citations;
inspect a custom graph's schema before describing any additional attributes.
An absent path within the requested hop limit is not evidence of no biological
relationship.

Personalized PageRank, permutation enrichment, and graph connections prioritize
associations and candidate mechanisms. They do not establish causality,
regulatory direction, or treatment effects. Use the result's enrichment
statistics, input measurements, and independently checked literature to assess
the hypotheses you identify.
