---
title: "Transcriptomic Knowledge Graph Integration with tKOI"
author: "Wanjun Gu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transcriptomic Knowledge Graph Integration with tKOI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`tKOI` (Transcriptomic Knowledge-graph Omics Integration) is an R package for integrating transcriptomic data with a biological knowledge graph to identify enriched biological concepts and pathways. It combines PageRank propagation with permutation testing and ontology-based annotations for high-resolution network interpretation.

# Step-by-Step Example

This vignette demonstrates a complete workflow using the `tkoi` package.

## Load Example Expression Data

```{r load-data, eval=FALSE}
file_path = system.file("extdata", "example_data.csv", package = "tkoi")
expression_data = data.table::fread(file_path)
head(expression_data)
```

## Run tKOI Analysis

```{r run-tkoi, eval=FALSE}
tkoi_result = run_tkoi(
  expression_data = expression_data,
  subnetwork = tkoi::tkoi_net,
  pvalue_threshold = 0.05,
  logfc_threshold = 0.25,
  indirect_link_threshold = 3,
  topology_similarity = 0.9,
  n_permutation = 100,
  damping_factor = 0.85,
  maximum_iteration = 500
)
```

## Perform Gene Ontology Enrichment

```{r enrichment, eval=FALSE}
tkoi_result = run_gene_enrichment(tkoi_result)
```

## Visualize GO vs Graph Enrichment

```{r go-plot1, eval=FALSE}
tkoi_result@gene_enrichment_comparison$comparison_scatter1
```

```{r go-plot2, eval=FALSE}
tkoi_result@gene_enrichment_comparison$comparison_scatter2
```

## Gene-Level Network Visualization

```{r gene-plot, eval=FALSE}
plt1 = make_gene_exploration_plot(
  tkoi_list = tkoi_result,
  sig_color = "#F39B7FB2",
  non_sig_color = "gray"
)
plt1
```

## Export Enrichment Summary Table

```{r export, eval=FALSE}
gene_data = export_gene_exploration_data(tkoi_result)
head(gene_data)
```

## Visualize Top Enriched Genes

```{r topn-plot, eval=FALSE}
plt2 = visualize_topn(
  tkoi_list = tkoi_result,
  category = "Gene",
  top_n = 25,
  high_color = "#FF5733",
  low_color = "#154360"
)
plt2
```

## Save Analysis Result (Optional)

```{r save-result, eval=FALSE}
save(tkoi_result, file = "tkoi_result.rda")
```

# Session Info

```{r session-info}
sessionInfo()
```
